---
title: "Bulk RNA-seq PCA"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

```{r, message = FALSE, warning = F}
library(ggbiplot)
library(tidyverse)
library(DESeq2)
# knitr::opts_chunk$set(cache = T, autodep = T)
```

```{r set_num_of_genes_for_pca}
num_of_genes.pca <- 1000
```

In this document, we are looking at the main components separating the samples, using PCA as an analyisis tool. We will use the top `r num_of_genes.pca` most variable genes.

Please refer to the [Data - Bulk RNAseq](data-bulkRNAseq.html) page for more info.

```{r read_data}
data("bulk4_dds")
data("paper_palette")
```

```{r function_plot_PCAs_from_vst}
plot_PCAs_from_vst <- function(vst, num.genes = 1000) {
  gene_variance <- rowVars(assay(vst))

  # select the 1000 top genes by variance
  selected_genes <- order(gene_variance, decreasing=TRUE) %>% head(n = num.genes)
  
  # perform a PCA on the data in assay(x) for the selected genes
  pca <- prcomp(t(assay(vst)[selected_genes, ]))
  
  pca_perc_variance <- round(1000 * pca$sdev^2 / sum(pca$sdev^2)) / 10
  
  pcaData <- pca$x[, 1:2] %>%
    as.data.frame() %>%
    rownames_to_column(var = "sample") %>%
    left_join(colData(vst) %>% as_tibble(), by = c("sample"))
  
  range_x <- range(pcaData$PC1)
  range_y <- range(pcaData$PC2)
  
  if (length(unique(pcaData$patient)) > 1) {
    ell_patients_groups <- pcaData %>% group_by(patient, group) %>%
      dplyr::summarise(n = n()) %>% filter(n > 2)
    ell <- list()
    apply(ell_patients_groups, 1, function(x) {
      this_patient = x[1]
      this_group = x[2]
      data <- pcaData %>% filter(patient == this_patient & group == this_group)
      this_ell <- car::dataEllipse(x = data$PC1, y = data$PC2,
                                   levels = 0.68, segments = 100,
                                   draw = F)
      ell[[paste(this_patient, this_group)]] <<-
        tibble(patient = this_patient, group = this_group,
               x = this_ell[1:100, 1], y = this_ell[1:100, 2],
             xend = this_ell[c(2:100, 1), 1], yend = this_ell[c(2:100, 1), 2])
      
    })
    ell <- Reduce(rbind, ell)

    g2 <- ggplot(pcaData, aes(PC1, PC2))
    if (length(ell) > 0) {
      g2 <- g2 +
        geom_segment(data = ell,
                     aes(x=x, y=y, xend=xend, yend=yend,
                         group=group, col=group, linetype = patient))
    }
    g2 <- g2 +
      geom_point(aes(col = group, shape = patient), size = 3) +
      # ggrepel::geom_text_repel(aes(label = sample), size = 2) +
      xlab(paste0("PC1: ", pca_perc_variance[1],"% variance")) +
      ylab(paste0("PC2: ", pca_perc_variance[2],"% variance")) +
      scale_colour_manual(values = paper_palette) +
      scale_fill_manual(values = paper_palette) +
      coord_fixed(ratio = diff(range_x) / diff(range_y)) +
      theme_bw(base_line_size = 0)
    g1 <- g2 +
      ggrepel::geom_text_repel(aes(label = sample), size = 2)

    ell_groups <- pcaData %>% group_by(group) %>%
      dplyr::summarise(n = n()) %>% filter(n > 2) %>% pull(group)
    ell <- lapply(ell_groups, function(name) {
      data <- pcaData %>% filter(group == name)
      car::dataEllipse(x = data$PC1, y = data$PC2,
                       levels = 0.68, segments = 100,
                       draw = F)
    })
    names(ell) <- ell_groups
    ell <- lapply(names(ell), function(name) {
      tibble(group = name, x = ell[[name]][1:100, 1], y = ell[[name]][1:100, 2],
             xend = ell[[name]][c(2:100, 1), 1], yend = ell[[name]][c(2:100, 1), 2])
      })
    ell <- Reduce(rbind, ell)
    
    g3 <- ggplot(pcaData, aes(PC1, PC2))
    if (length(ell) > 0) {
      g3 <- g3 +
        geom_segment(data = ell,
                     aes(x = x, y = y, xend = xend, yend = yend,
                         group = group, col = group))
      
    }
    g3 <- g3 +
      stat_ellipse(aes(col = group), level = 0.68, type = "norm") +
      geom_point(aes(col = group, shape = patient), size = 3) +
      # ggrepel::geom_text_repel(aes(label = sample), size = 2) +
      xlab(paste0("PC1: ", pca_perc_variance[1],"% variance")) +
      ylab(paste0("PC2: ", pca_perc_variance[2],"% variance")) +
      scale_colour_manual(values = paper_palette) +
      scale_fill_manual(values = paper_palette) +
      coord_fixed(ratio = diff(range_x) / diff(range_y)) +
      theme_bw(base_line_size = 0)
    
    print(g1)
    print(g2)
    print(g3)
    
    return(invisible(list(pcaData, g1, g2, g3)))
    
  } else {
    ell_groups <- pcaData %>% group_by(group) %>%
      dplyr::summarise(n = n()) %>% filter(n > 2) %>% pull(group)
    ell <- lapply(ell_groups, function(name) {
      data <- pcaData %>% filter(group == name)
      car::dataEllipse(x = data$PC1, y = data$PC2,
                       levels = 0.68, segments = 100,
                       draw = F)
    })
    names(ell) <- ell_groups
    ell <- lapply(names(ell), function(name) {
      tibble(group = name, x = ell[[name]][1:100, 1], y = ell[[name]][1:100, 2],
             xend = ell[[name]][c(2:100, 1), 1], yend = ell[[name]][c(2:100, 1), 2])
      })
    ell <- Reduce(rbind, ell)

    g2 <- ggplot(pcaData, aes(PC1, PC2))
    if (length(ell) > 0) {
      g2 <- g2 +
        geom_segment(data = ell,
                     aes(x = x, y = y, xend = xend, yend = yend,
                         group = group, col = group))
      
    }
    g2 <- g2 +
      geom_point(aes(fill = group), shape = 21, color = "black", size = 3) +
      xlab(paste0("PC1: ", pca_perc_variance[1],"% variance")) +
      ylab(paste0("PC2: ", pca_perc_variance[2],"% variance")) +
      scale_colour_manual(values = paper_palette) +
      scale_fill_manual(values = paper_palette) +
      coord_fixed(ratio = diff(range_x) / diff(range_y)) +
      theme_bw(base_line_size = 0)
    g1 <- g2 +
      ggrepel::geom_text_repel(aes(label = sample), size = 2)

    print(g1)
    print(g2)

    return(invisible(list(pcaData, g1, g2)))
  
  }
  
}
```


We transform the data all together with the VST transformation function. This essentially log-transforms the counts while taking into account the library size (see DESeq2 manual for more info).

```{r vst_transform}
bulk4_vst <- vst(bulk4_dds)
```


## Treated vs untreated samples

### All samples

This is the initial PCA, built from the `bulk4_dds` data set, using VST-transformed counts.

```{r treated_vs_untreated}
this_bulk_vst <- bulk4_vst[, colData(bulk4_dds)$patient %in% c("PT1", "PT12", "PT13")]
this_bulk_vst$patient <- factor(c("Pt1", "Pt2", "Pt12", "Pt13")[as.numeric(this_bulk_vst$patient)])
plots <- plot_PCAs_from_vst(this_bulk_vst)
```

#### Extended Figure 5a

<div class="alert alert-info">The next chunk of code save the ExtFig5a figure and associated data into the `output` folder</div>

```{r save_extfig_5a}
plots[[4]] +
  coord_cartesian() +
  guides(color = "none") +
  guides(shape = guide_legend(title = "Patient")) +
  theme(legend.position = "top")
ggsave("output/figures/ExtFig5a_pca_3patients.pdf",
       device = "pdf", width = 5, height = 4, scale = 0.8)

excel_data <- plots[[1]]

wb <- openxlsx::createWorkbook()
ws <- openxlsx::addWorksheet(wb, "ExtFig5a PCA bulk RNAseq")

openxlsx::writeDataTable(wb, sheet = ws, x = excel_data,
               rowNames = F, tableStyle = "none", withFilter = F)
openxlsx::setColWidths(wb, ws, cols = 1:ncol(excel_data), widths = "auto")
openxlsx::addStyle(wb, ws, rows = 1, cols = 1:ncol(excel_data),
                   style = openxlsx::createStyle(textDecoration = "bold"))
openxlsx::addStyle(wb, ws, rows = 1:nrow(excel_data) + 1, cols = 1,
                   style = openxlsx::createStyle(textDecoration = "bold"))

openxlsx::saveWorkbook(wb, "output/tables/ExtFig5a_bulkRNAseq_data.xlsx", overwrite = TRUE)
```

### PT1 only

Same as above, but using samples from PT1 only.

```{r treated_vs_untreated.pt1}
this_bulk_vst <- bulk4_vst[, colData(bulk4_dds)$patient %in% c("PT1")]
plot_PCAs_from_vst(this_bulk_vst)
```


### PT12 only

Same as above, but using samples from PT12 only.

```{r treated_vs_untreated.pt12}
this_bulk_vst <- bulk4_vst[, colData(bulk4_dds)$patient %in% c("PT12")]
plot_PCAs_from_vst(this_bulk_vst)
```

### PT13 only

Same as above, but using samples from PT13 only.

```{r treated_vs_untreated.pt13}
this_bulk_vst <- bulk4_vst[, colData(bulk4_dds)$patient %in% c("PT13")]
plot_PCAs_from_vst(this_bulk_vst)
```

## Treatment response experiment

This is the initial PCA, built from the `bulk4_dds` data set, using VST-transformed counts. The data correspond to PT2, where we have five types of samples:

* untreated
* treatment withdrawn
* chronically treated
* acutely treated
* relapse (let to relapse)

```{r treatment_response}
this_bulk_vst <- bulk4_vst[, colData(bulk4_dds)$patient %in% c("PT2")]
hack_bulk_vst <- this_bulk_vst
hack_bulk_vst$patient <- hack_bulk_vst$tissue
plots <- plot_PCAs_from_vst(hack_bulk_vst)
```

#### Extended Figure 5b

<div class="alert alert-info">The next chunk of code save the ExtFig5b figure and associated data into the `output` folder</div>

```{r save_extfig_5b}
plots[[4]] +
  coord_cartesian() +
  guides(color = "none") +
  guides(fill = "none") +
  guides(shape = guide_legend(title = "Tissue")) +
  theme(legend.position = "top")
ggsave("output/figures/ExtFig5b_pca_treatment_response.pdf",
       device = "pdf", width = 6, height = 4, scale = 0.8)

excel_data <- plots[[1]]

wb <- openxlsx::createWorkbook()
ws <- openxlsx::addWorksheet(wb, "ExtFig5b PCA bulk RNAseq")

openxlsx::writeDataTable(wb, sheet = ws, x = excel_data,
               rowNames = F, tableStyle = "none", withFilter = F)
openxlsx::setColWidths(wb, ws, cols = 1:ncol(excel_data), widths = "auto")
openxlsx::addStyle(wb, ws, rows = 1, cols = 1:ncol(excel_data),
                   style = openxlsx::createStyle(textDecoration = "bold"))
openxlsx::addStyle(wb, ws, rows = 1:nrow(excel_data) + 1, cols = 1,
                   style = openxlsx::createStyle(textDecoration = "bold"))

openxlsx::saveWorkbook(wb, "output/tables/ExtFig5b_bulkRNAseq_data.xlsx", overwrite = TRUE)
```

