---
title: "Data - Bulk RNAseq"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r, message = F, warning = F}
library(tidyverse)
library(DESeq2)
library(DT)
# knitr::opts_chunk$set(cache = T, autodep = T)
```


This is based on merged_gene_counts.bulk.txt.gz and includes all four patients (PT1, PT2, PT12 and PT13). PT12 and PT13 were added at a later date and the data were re-processed all together from scratch with the NextFlow pipeline described above.

## Initial filtering

In this section, we remove:

* genes that are not expressed in almost all samples. More precisely, we remove any gene that does not have at least 1 read count in at least 3 samples.
* samples that do not have at least 20 genes expressed (min read counts > 0)

Then we plot several PCA plots to identify outliers.

```{r data_wrangling}
merged_gene_counts.bulk <- read.table("data-raw/merged_gene_counts.bulk.txt.gz", row.names = 1, header = T)
colnames(merged_gene_counts.bulk) <- gsub(".", "_", colnames(merged_gene_counts.bulk), fixed = T)
sample_table <- read.csv("data-raw/sample_bulk_rna.csv", stringsAsFactors = F)


# Identify genes that are not expressed in any sample
minExpression <- 1
minSamples <- 3
genes_in_few_samples <- names(which(apply(merged_gene_counts.bulk >= minExpression, 1, sum) >= minSamples))
merged_gene_counts.bulk <- merged_gene_counts.bulk[genes_in_few_samples, ]

# Identify samples that don't have at least 20 genes expressed (very low hanging fruit)
minExpression <- 1
minGenes <- 20
samples_with_few_genes <- names(which(apply(merged_gene_counts.bulk >= minExpression, 2, sum) < minGenes))
samples_to_remove <- c(samples_with_few_genes)

# Filter sample table
sample_table <- sample_table %>% filter(!(title %in% samples_to_remove))

# Filter counts matrix
merged_gene_counts.bulk <- merged_gene_counts.bulk[, sample_table$title]

# Data wrangling, from EGA metadata
sample_table <- sample_table %>%
  dplyr::rename(sample = title, patient = subjectId, group = description) %>%
  dplyr::select(sample, patient, group) %>%
  dplyr::mutate(tissue = case_when(
    patient == "PT1" ~ "",
    patient == "PT12" ~ "",
    patient == "PT13" ~ "",
    grepl("mouse bone marrow", group) ~ "BM",
    grepl("mouse spleen", group) ~ "Spleen",
    grepl("mouse brain", group) ~ "Brain",
    TRUE ~ group
  )) %>%
  dplyr::mutate(group = case_when(
    patient %in% c("PT1", "PT12", "PT13") & grepl("untreated", group) ~ "Untreated",
    patient %in% c("PT1", "PT12", "PT13") & grepl("treated", group) ~ "Treated",
    patient == "PT2" & grepl("untreated control", group) ~ "Never treated",
    patient == "PT2" & grepl("acutely treated", group) ~ "Acutely treated",
    patient == "PT2" & grepl("treatment withdr", group) ~ "Treatment withdrawn",
    patient == "PT2" & grepl("chronically treated", group) ~ "Chronically treated",
    patient == "PT2" & grepl("relapse", group) ~ "Relapse",
    TRUE ~ group
  )) %>%
  dplyr::mutate(patient = factor(patient), group = factor(group), tissue = factor(tissue))

bulk_dds <- DESeq2::DESeqDataSetFromMatrix(merged_gene_counts.bulk,
                                           colData = sample_table,
                                           design = ~ group)

bulk_vst <- DESeq2::vst(bulk_dds)
pca.pre1 <- DESeq2::plotPCA(bulk_vst[, colData(bulk_vst)$patient == "PT1"], intgroup = c("patient", "group")) +
  ggrepel::geom_text_repel(aes(label = name), size = 2) +
  ggtitle("PT1")
pca.pre2 <- DESeq2::plotPCA(bulk_vst[, colData(bulk_vst)$patient == "PT2"], intgroup = c("patient", "group")) +
  ggrepel::geom_text_repel(aes(label = name), size = 2) +
  ggtitle("PT2")
pca.pre3 <- DESeq2::plotPCA(bulk_vst[, colData(bulk_vst)$patient == "PT12"], intgroup = c("patient", "group")) +
  ggrepel::geom_text_repel(aes(label = name), size = 2) +
  ggtitle("PT12")
pca.pre4 <- DESeq2::plotPCA(bulk_vst[, colData(bulk_vst)$patient == "PT13"], intgroup = c("patient", "group")) +
  ggrepel::geom_text_repel(aes(label = name), size = 2) +
  ggtitle("PT13")
pca.pre5 <- DESeq2::plotPCA(bulk_vst[, colData(bulk_vst)$patient %in% c("PT1", "PT12")], intgroup = c("patient", "group")) +
  ggrepel::geom_text_repel(aes(label = name), size = 2) +
  ggtitle("PT1 + PT12")
pca.pre6 <- DESeq2::plotPCA(bulk_vst[, colData(bulk_vst)$patient %in% c("PT1", "PT12", "PT13")], intgroup = c("patient", "group")) +
  ggrepel::geom_text_repel(aes(label = name), size = 2) +
  ggtitle("PT1 + PT12 + PT13")
pca.pre7 <- DESeq2::plotPCA(bulk_vst, intgroup = c("patient", "group")) +
  ggrepel::geom_text_repel(aes(label = name), size = 2) +
  ggtitle("All patients")

pca.pre1
pca.pre2
pca.pre3
pca.pre4
pca.pre5
pca.pre6
pca.pre7
```

## Removing the outliers

List of outliers:

```{r removing_outliers}
samples_to_remove <- c("VT11_N705_S504", # Outlier in PT1
                       "VTb2_N706_S517", # Acutely treated clustering with untreated
                       "VT47b_N706_S503" # Outlier in PT12
                       )

sample_table %>%
  filter(sample %in% samples_to_remove) %>%
  datatable(rownames = F, options = list(pageLength = 100, ordering = F, dom = 't'))
  
sample_table <- sample_table %>% filter(!(sample %in% samples_to_remove))

merged_gene_counts.bulk <- merged_gene_counts.bulk[, sample_table$sample]

bulk_dds <- DESeq2::DESeqDataSetFromMatrix(merged_gene_counts.bulk,
                                           colData = sample_table,
                                           design = ~ group)

bulk_vst <- DESeq2::vst(bulk_dds)

pca.post1 <- DESeq2::plotPCA(bulk_vst[, colData(bulk_vst)$patient == "PT1"], intgroup = c("patient", "group")) +
  ggrepel::geom_text_repel(aes(label = name), size = 2) +
  ggtitle("PT1")
pca.post2 <- DESeq2::plotPCA(bulk_vst[, colData(bulk_vst)$patient == "PT2"], intgroup = c("patient", "group")) +
  ggrepel::geom_text_repel(aes(label = name), size = 2) +
  ggtitle("PT2")
pca.post3 <- DESeq2::plotPCA(bulk_vst[, colData(bulk_vst)$patient == "PT12"], intgroup = c("patient", "group")) +
  ggrepel::geom_text_repel(aes(label = name), size = 2) +
  ggtitle("PT12")
pca.post4 <- DESeq2::plotPCA(bulk_vst[, colData(bulk_vst)$patient == "PT13"], intgroup = c("patient", "group")) +
  ggrepel::geom_text_repel(aes(label = name), size = 2) +
  ggtitle("PT13")
pca.post5 <- DESeq2::plotPCA(bulk_vst[, colData(bulk_vst)$patient %in% c("PT1", "PT12")], intgroup = c("patient", "group")) +
  ggrepel::geom_text_repel(aes(label = name), size = 2) +
  ggtitle("PT1 + PT12")
pca.post6 <- DESeq2::plotPCA(bulk_vst[, colData(bulk_vst)$patient %in% c("PT1", "PT12", "PT13")], intgroup = c("patient", "group")) +
  ggrepel::geom_text_repel(aes(label = name), size = 2) +
  ggtitle("PT1 + PT12 + PT13")
pca.post7 <- DESeq2::plotPCA(bulk_vst, intgroup = c("patient", "group")) +
  ggrepel::geom_text_repel(aes(label = name), size = 2) +
  ggtitle("All patients")
```

And here are the PCA plots after and before removing the outliers:

```{r comparison_before_after_outliers, results='asis'}
cat("### {.tabset .unlisted .unnumbered .toc-ignore}\n\n")
cat("#### Before\n\n")
pca.pre1
cat("#### After {.active}\n\n")
pca.post1

cat("### {.tabset .unlisted .unnumbered .toc-ignore}\n\n")
cat("#### Before\n\n")
pca.pre2
cat("#### After {.active}\n\n")
pca.post2

cat("### {.tabset .unlisted .unnumbered .toc-ignore}\n\n")
cat("#### Before\n\n")
pca.pre3
cat("#### After {.active}\n\n")
pca.post3

cat("### {.tabset .unlisted .unnumbered .toc-ignore}\n\n")
cat("#### Before\n\n")
pca.pre4
cat("#### After {.active}\n\n")
pca.post4

cat("### {.tabset .unlisted .unnumbered .toc-ignore}\n\n")
cat("#### Before\n\n")
pca.pre5
cat("#### After {.active}\n\n")
pca.post5

cat("### {.tabset .unlisted .unnumbered .toc-ignore}\n\n")
cat("#### Before\n\n")
pca.pre6
cat("#### After {.active}\n\n")
pca.post6

cat("### {.tabset .unlisted .unnumbered .toc-ignore}\n\n")
cat("#### Before\n\n")
pca.pre7
cat("#### After {.active}\n\n")
pca.post7
```


## Storing the data

The data objects stored are called: `bulk4_counts` and `bulk4_dds`.

```{r}
# Save this data object
usethis::use_directory("data")
bulk4_counts <- merged_gene_counts.bulk
save(bulk4_counts, file = "data/bulk4_counts.rda")

usethis::use_directory("data")
bulk4_dds <- bulk_dds
save(bulk4_dds, file = "data/bulk4_dds.rda")
```

