---
title: "Data"
output:
  workflowr::wflow_html:
    toc: false
editor_options:
  chunk_output_type: console
---

<div class="alert alert-warning">Under construction</div>

```{r, message = FALSE, warning = F}
library(tidyverse)
library(DT)
```

#### 1. Samples overview

This table show the number of datasets per sample per patient. The numbers correspond to the number of cells or samples **after QC**.

```{r samples_summary}
samples_summary <- read_csv("data-raw/samples_summary.csv", col_types = "cccc")

samples_summary %>%
  pivot_wider(names_from = "Assay", values_from = n_samples, values_fill = list(n_samples = "")) %>%
  select(Patient, Sample, `Bulk RNA`, scRNA, scWGS, EPIC, `Bulk WGS`, `sc Q-PCR`) %>%
  cbind(prev_patient = c("", .$Patient[1:(nrow(.)-1)])) %>%
  mutate(Patient = ifelse(Patient == prev_patient, "", Patient)) %>%
  select(-prev_patient) %>%
  datatable(rownames = F, options = list(pageLength = 100, ordering = F, dom = 't')) %>%
  formatStyle(c('Patient', 'Sample'), fontWeight = "bold") %>%
  formatStyle(2:8, 'Sample', backgroundColor =
                styleEqual(c("Diagnosis", "MRD (G&T)", "Relapse",
                             "PDX untreated", "PDX treated", "PDX withdrawn",
                             "PDX chronically", "PDX acutely", "PDX relapse",
                             "chord blood"),
                           c("#FCE4D6", "#FCE4D6", "#FCE4D6",
                             "#DDEBF7", "#DDEBF7", "#DDEBF7",
                             "#DDEBF7", "#DDEBF7", "#DDEBF7",
                             "#E2EFDA"))) %>%
  formatStyle(3:8, backgroundColor = styleEqual("", "#D9D9D9")) %>%
  formatStyle(1:8, borderColor = "black") %>%
  formatStyle(1:8, "Patient", borderColor = styleEqual("", "#CCCCCC")) %>%
  formatStyle(1, borderLeft = "solid 1px") %>%
  formatStyle(c(1, 2, 8), borderRight = "solid 1px") %>%
  formatStyle(c(3:7), borderRight = "solid 1px #CCCCCC")
```


#### 2. [Bulk RNA-seq](data-bulkRNAseq.html)

Preprocessing of the bulk RNA-seq data

#### 3. scRNA-seq primaries

*Under construction*

#### 4. scRNA-seq PDX

*Under construction*

#### 5. [Ancillary data](data-ancillary.html)

Gene sets and color palette.
