---
title: "Ancillary Data"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, warning = FALSE, message = FALSE}
# knitr::opts_chunk$set(cache = T, autodep = T)

library(tidyverse)
library(DT)
```

## Signatures and gene sets

Several signatures/gene sets are used in this analysis.

* **MSigDB**: all signatures have been downloaded using the `msigdbr` package.

* **Lineage signatures**: from Laurenti et al (ProB, MLP, MEP, HSC, GMP, ETP, EarlyB, CMP)

* **Cell state**: from Tirosh et al (Tirosh_G1_S, Tirosh_G2_M, TIROSH_Stemness)


```{r}
msigdbr_full_signatures_table <- msigdbr::msigdbr("Homo sapiens")

groups <- msigdbr_full_signatures_table %>%
  group_by(gs_cat, gs_subcat) %>%
  summarise(n = n()) %>%
  mutate(name = ifelse(gs_subcat != "", paste0(gs_cat, "-", gs_subcat), gs_cat))

signatures <- list()
sink <- groups %>% apply(1, function(x) {
  signatures[[x["name"]]] <<- msigdbr_full_signatures_table %>%
    filter(gs_cat == x["gs_cat"] & gs_subcat == x["gs_subcat"]) %>%
    split(x = .$gene_symbol, f = .$gs_name)
  })

# This table is available from Laurenti et al.
laurenti_signatures <- read.table("data-raw/PopSpecificSignatures_250.gmx", header = T)

sink <- lapply(colnames(laurenti_signatures), function(x) {
  signatures[["lineages"]][[x]] <<- as.character(laurenti_signatures[, x])
  })

# This table was extracted from Tirosh et al.
cell_state_signatures <- fgsea::gmtPathways("data-raw/cell_state.gmt")

sink <- lapply(names(cell_state_signatures), function(x) {
  signatures[["cell_state"]][[x]] <<- as.character(cell_state_signatures[[x]])
  })


usethis::use_directory("data")

save(signatures, file = "data/signatures.rda")
```

Here is a list of the different signature sets:

```{r}
enframe(sapply(signatures, length), name = "set", "number") %>%
  left_join(enframe(sapply(signatures, function(x) {
    num <- length(x)
    if (num > 10) {
      names <- paste(c(names(x)[1:10], "..."), collapse = ",")
    } else {
      names <- paste(names(x), collapse = ",")
    }
    return(names)
    }), name = "set", "signatures"), by = "set") %>%
  DT::datatable()
```

To use the signatures, simply do:

```
data("signatures")

hallmarks <- signatures[["H"]]
lineages <- signatures[["lineages"]]
cell_cycle_genes <- signatures[["lineages"]][["GO_CELL_CYCLE"]]

# Combinations
pathways <- c(signatures[["lineages"]], signatures[["H"]])
names(pathways) <- gsub("HALLMARK_", "", names(pathways))
```


## Paper Palette

To use the paper palette with ggplot2, simply do:

```
data("paper_palette")

ggplot() +
  ... +
  scale_fill_manual(values = paper_palette) +
  ...
```

```{r}
paper_palette <- c(
  ## Bulk treated
  "BulkChemotherapy" = "#AF4FAF",
  "Chemotherapy" = "#AF4FAF",
  "Treated" = "#AF4FAF",
  
  ## Bulk relapse (240, 146, 242)
  "BulkRelapse" = "#F092F2",
  "Relapse" = "#F092F2",
  
  ## Bulk untreated (for simple)
  "BulkUntreated" = "#43B0E3",
  "Untreated" = "#43B0E3",

  ## Diagnosis
  "Diagnosis" = "#E38522",
  "MRD" = "#30AF30",

  ## PRS
  "PRS" = "#307F7F",
  
  ## Cell cycle: deep quiescent
  "Deep" = "#959867",
  "DeepQ" = "#959867",
  
  ## Cell cycle: shallow quiescent
  "Shallow" = "#94BE69",
  "ShallowQ" = "#94BE69",
  
  ## Cell cycle: cycling
  "Cycling" = "#CA4B1F",
  "Cycling" = "#CA4B1F",
  
  ## Bulk headache: T/T (231, 119, 108)
  "chem-chem" = "#E7776C",
  "Chronically treated" = "#E7776C",
  
  ## Bulk headache: T/U
  "chem-nochem" = "#A1A22C",
  "Treatment Withdrawn" = "#A1A22C",
  "Treatment withdrawn" = "#A1A22C",
  
  ## Bulk headache: U/T
  "nochem-chem" = "#55B583",
  "Acutely treated" = "#55B583",
  
  ## Bulk headache: U/U
  "nochem-nochem" = "#0416C8",
  "Untreated" = "#0416C8",
  "Never treated" = "#0416C8",

  ## Lineages
  "ProB" = "#AF4AFF",
  "MLP" = "#797FF4",
  "MEP" = "#44B4FF",
  "HSC" = "#4BD6B3",
  "GMP" = "#67F16C",
  "ETP" = "#97A645",
  "EarlyB" = "#CCB43D",
  "CMP" = "#F88134",
  "pre-BII" = "#FF4C24",
  "pre-BI" = "#AB0300",
  
  ## Yes/No
  "Yes"= "#307F7F",
  "No" = "#B4B4B4")

usethis::use_directory("data")
save(paper_palette, file = "data/paper_palette.rda")
```

