---
title: "Pt2 diagnosis resampling analysis"
author: "Javier Herrero"
date: "`r date()`"
output:
  html_document:
    code_folding: hide
    toc: TRUE
    toc_float: TRUE
---

```{r}
knitr::read_chunk("resampling_chunks.R")
```

```{r setup, include=FALSE}
library(tidyverse)
library(cowplot)
library(ggExtra)
library(knitr)
library(optrees)
library(mclust)
library(DT)

knitr::opts_chunk$set(echo = TRUE)
opts_chunk$set(cache=TRUE)
opts_chunk$set(cache.lazy=FALSE)
opts_chunk$set(autodep = TRUE)

source("plots.R")
source("cnv_analysis.lib.R")
source("resampling_functions.R")
```

```{r set_seed}
set.seed(123456)
```

# Input data {.tabset}

```{r read_input, results = "asis"}
single_seg_renorm <- readRDS("cnv_copynumber_Pt2_dia_100k.single_renorm.RDS")
multi_seg_renorm <- readRDS("cnv_copynumber_Pt2_dia_100k.multi_seg_renorm.RDS")
library(xlsx)
meta_data <- read.xlsx(paste0("../data/Samples.xlsx"), sheetIndex = 1)
cells <- meta_data %>% filter(Day == "Diagnosis" & Patient == "Pt2") %>% pull(Cell) %>% as.character()
cells <- intersect(colnames(multi_seg_renorm[["05"]][, -c(1:3)]), cells)
multi_seg_renorm[["05"]] <- multi_seg_renorm[["05"]][, c("CHR", "START", "END", cells)]
single_seg_renorm[["10"]] <- single_seg_renorm[["10"]][, c("CHR", "START", "END", cells)]
```

There are `r length(cells)` cells in this dataset.

```{r define_cnvs, results = "asis"}
plotGenome(multi_seg_renorm[['05']])
plotGenome(single_seg_renorm[['10']])
plotFrequency(multi_seg_renorm[['05']])
plotFrequency(single_seg_renorm[['10']])
for (this_chr in paste0("chr", c(1:22, "X"))) {
  if (!is.null(opts_knit$get("output.dir"))) {
    cat(knit_child(text = paste0("\n## ", this_chr, "\n\n"), quiet = T))
  }
  plotChr(multi_seg_renorm[['05']], this_chr)
  plotChr(single_seg_renorm[['10']], this_chr)
}

get_cnv_table <- function(data) {
  cnv_table <- tibble()
  all_cells <- colnames(data)[-c(1:3)]
  chr_length_rle <- rle(data[, "CHR"])
  chr_starts <- cumsum(chr_length_rle$lengths) - chr_length_rle$lengths + 1
  names(chr_starts) <- chr_length_rle$values
  for (this_cell in all_cells) {
    for (this_chr in unique(data[, "CHR"])) {
      this_data <- data[data$CHR == this_chr, this_cell]
      rle <- rle(this_data)
      this_cnv_table <- tibble(
        from = cumsum(rle$lengths) - rle$lengths + chr_starts[this_chr],
        to = cumsum(rle$lengths) + chr_starts[this_chr] - 1,
        n = rle$values,
        num.cells = 1,
        cells = this_cell,
        chr = data[cumsum(rle$lengths) - rle$lengths + chr_starts[this_chr], "CHR"],
        start = data[cumsum(rle$lengths) - rle$lengths + chr_starts[this_chr], "START"],
        end = data[cumsum(rle$lengths) +  chr_starts[this_chr] - 1, "END"],
        n.bins = rle$lengths
        )
      cnv_table <- rbind(cnv_table, this_cnv_table %>% filter(n != 2))
    }
  }
  cnv_table <- cnv_table %>%
    group_by(from, to, n, chr, start, end, n.bins) %>%
    summarise(num.cells = n(), cells = paste(cells, collapse = ","))
  return(cnv_table)
}

row.names(multi_seg_renorm[['05']]) <- NULL
cnv_table <- get_cnv_table(multi_seg_renorm[['05']])
# cnv_table_orig <- cnv_table
# cnv_table <- cnv_table_orig
cnv_table <- cnv_table %>% filter(n.bins > 6)
cnv_table <- cnv_table %>% filter(chr != "chrY")

to_be_removed <- c(1,2,3,4,10,12,13,16,17,18,19,21,22,24,25,26,28,29,30,35,36,37,38,39)
not_so_sure <- c()
to_be_kept <- c(5,6,7,8,9,11,12,14,15,20,23,27,31,32,33,34)
curated_cnv_table <- as.data.frame(cnv_table)[to_be_kept, ]

curated_cnv_table["7", "start"] <- 64234606
curated_cnv_table["7", "end"] <- 66386004
curated_cnv_table["7", "n"] <- 3
curated_cnv_table["14", "start"] <- 35109012
curated_cnv_table["14", "end"] <- 37934244
curated_cnv_table["14", "n"] <- 3
curated_cnv_table["15", "end"] <- 38841621
curated_cnv_table["31", "end"] <- 18959018
curated_cnv_table["33", "end"] <- 23660545
curated_cnv_table["34", "end"] <- 155049231





cnv_table <- as.data.frame(curated_cnv_table)

#### --- Attempt at using single_sample segmentation ----
# library(GenomicRanges)
# row.names(single_seg_renorm[['10']]) <- NULL
# single.cnv_table <- get_cnv_table(single_seg_renorm[['10']])
# single.cnv_table <- single.cnv_table %>% filter(n.bins > 6)
# single.cnv_table <- single.cnv_table %>% filter(chr != "chrY")
# 
# single.cnv_table.gr <- GenomicRanges::GRanges(
#   seqnames = single.cnv_table$chr,
#   ranges = IRanges::IRanges(start = single.cnv_table$start,
#                             end = single.cnv_table$end),
#   n = single.cnv_table$n
# )
# single.cnv_table.grl <- split(single.cnv_table.gr, single.cnv_table.gr$n)
# single.cnv_table.grl.ranges <- reduce(single.cnv_table.grl)
# collated.single.cnv_table <- single.cnv_table[0, ]
# for (this_n in names(single.cnv_table.grl)) {
#   overlaps <- findOverlaps(single.cnv_table.grl.ranges[[this_n]],
#                            single.cnv_table.grl[[this_n]])
#   ranges <- as.data.frame(single.cnv_table.grl.ranges[[this_n]]) %>% as_tibble() %>%
#     transmute(chr = seqnames, start = start, end = end, range_id = row_number())
#   cnvs <- single.cnv_table %>% filter(n == this_n) %>% ungroup() %>% mutate(cnv_id = row_number())
#   new_cnvs <-     ranges %>%
#     left_join(overlaps %>% as_tibble(), by = c("range_id" = "queryHits")) %>%
#     left_join(cnvs, by = c("subjectHits" = "cnv_id")) %>%
#     group_by(range_id) %>% summarise(from = min(from), to = max(to), n = min(n),
#       chr = dplyr::first(chr.x), start = min(start.y), end = max(end.y), n.bins = 0,
#       num.cells = length(unique(unlist(strsplit(cells, ",")))),
#       cells = paste(unique(unlist(strsplit(cells, ","))), collapse = ",")) %>%
#     mutate(n.bins = to - from + 1) %>%
#     select(-range_id)
# 
#   if (nrow(collated.single.cnv_table) > 0) {
#     collated.single.cnv_table <- collated.single.cnv_table %>% rbind(new_cnvs)
#   } else {
#     collated.single.cnv_table <- new_cnvs
#   }
# }
# 
# all_cells <- colnames(single_seg_renorm[['10']])[-c(1:3)]
# collated.single.cnv_table <- collated.single.cnv_table %>% filter(
#     ((end - start) > 0000000 & num.cells > ceiling(length(all_cells) * 0.10)) |
#     ((end - start) > 1000000 & num.cells > ceiling(length(all_cells) * 0.05)) |
#     ((end - start) > 3000000 & num.cells > ceiling(length(all_cells) * 0.01)) |
#     ((end - start) > 5000000 & num.cells > ceiling(length(all_cells) * 0.00))
#   )
# cnv_table <- collated.single.cnv_table
#### --- Attempt at using single_sample segmentation ----

# cnv_table <- cnv_table %>% filter(chr != "chr2")
# cnv_table <- cnv_table %>% filter(chr != "chr3")
# cnv_table <- cnv_table %>% filter(chr != "chr4")
# # cnv_table <- cnv_table %>% filter(chr != "chr5")
# cnv_table <- cnv_table %>% filter(chr != "chr6")
# cnv_table <- cnv_table %>% filter(chr != "chr7")
# # cnv_table <- cnv_table %>% filter(chr != "chr8")
# # cnv_table <- cnv_table %>% filter(chr != "chr9")
# cnv_table <- cnv_table %>% filter(chr != "chr10")
# cnv_table <- cnv_table %>% filter(chr != "chr11")
# cnv_table <- cnv_table %>% filter(chr != "chr12")
# cnv_table <- cnv_table %>% filter(chr != "chr13")
# cnv_table <- cnv_table %>% filter(chr != "chr14")
# # cnv_table <- cnv_table %>% filter(chr != "chr15")
# cnv_table <- cnv_table %>% filter(chr != "chr16")
# cnv_table <- cnv_table %>% filter(chr != "chr17")
# cnv_table <- cnv_table %>% filter(chr != "chr18")
# # cnv_table <- cnv_table %>% filter(chr != "chr19")
# cnv_table <- cnv_table %>% filter(chr != "chr20")
# # cnv_table <- cnv_table %>% filter(chr != "chr21")
# # cnv_table <- cnv_table %>% filter(chr != "chr22")
# # cnv_table <- cnv_table %>% filter(chr != "chrX")
# 
# cnv_table <- cnv_table %>% filter(chr != "chr9" | end != 25222566)
# cnv_table <- cnv_table %>% filter(chr != "chr9" | start != 20620222 | end != 22976748)
# cnv_table <- cnv_table %>% filter(chr != "chr9" | start != 20620222 | end != 22774450 | n != 1)
# cnv_table <- cnv_table %>% filter(chr != "chr9" | start != 25325169 | end != 38841621)
# 
# cnv_table <- cnv_table %>% filter(chr != "chr15" | start != 20226353)
# cnv_table <- cnv_table %>% filter(chr != "chr15" | end < 23718060)
# 
# cnv_table <- cnv_table %>% filter(chr != "chr19" | start != 1)
# 
# cnv_table <- cnv_table %>% filter(chr != "chr21" | end != 21000898)
# 
# cnv_table <- cnv_table %>% filter(chr != "chrX" | (start == 2583625 & end == 155049231))
```

```{r get_signal_matrix}
data_env <- new.env()
load("data_Pt2_dia_100k.rda", envir = data_env)
segrenorm_data <- data_env$segrenorm_data
rm(data_env)
segrenorm_data <- segrenorm_data[, c("CHR", "START", "END", cells)]

# ## Adjust for ploidy:
# # 1. read ploidy data (from Ginkgo)
# ploidy <- read.table("../data/100k/ploidy.txt", header = T, row.names = 1)
# # 2. Multiply signal (ca.1 centered) by ploidy
# segrenorm_data[, -c(1:3)] <- sweep(segrenorm_data[, -c(1:3)], 2,
#                                    ploidy[colnames(segrenorm_data)[-c(1:3)], ], "*")
# # 2. Divide signal by 2 (to make it 1 centered again, after ploidy adjustment)
# segrenorm_data[, -c(1:3)] <- sweep(segrenorm_data[, -c(1:3)], 2, 2, "/")

old_signal_matrix <- matrix(0, nrow = ncol(segrenorm_data) - 3, ncol = nrow(cnv_table))
colnames(old_signal_matrix) <- paste0("V", 1:ncol(old_signal_matrix))
for (i in 1:nrow(cnv_table)) {
  this_cnv <- cnv_table[i, , drop = T]

  old_signal_matrix[, i] <- segrenorm_data %>%
    filter(CHR == this_cnv$chr & START >= this_cnv$start & END <= this_cnv$end) %>%
    select(-c(1:3)) %>%
    colMeans() * 2
}
cns <- cnv_table$n
```


# Curated CN events before resampling {.tabset}

```{r set_starting_thresholds, dependson = c("get_signal_matrix")}
starting_thresholds <- list(
  mean = ifelse(cns < 2, cns + 0.5, cns - 0.5),
  sd = rep(0.05, length(cns)))
starting_thresholds$mean[8] <- 1.5
```

```{r, cache = F}
DT::datatable(cnv_table)
```

```{r no-initial_density_plots, results = "asis", cache.extra = file.info("resampling_chunks.R")$mtime, dependson=c("define_cnvs", "get_signal_matrix", "set_starting_thresholds")}
```

```{r cache=F}
# knitr::knit_exit()
```


<!--
# Scatter plot

```{r no-pairwise_scatterplots, results = "asis", cache.extra = file.info("resampling_chunks.R")$mtime, dependson=c("define_cnvs", "get_signal_matrix", "set_starting_thresholds")}
```
-->

```{r cache=F}
# knitr::knit_exit()
```


# Reconstruct the tree

## Naive tree

```{r naive_tree, dependson=c("set_seed", "define_cnvs", "get_signal_matrix", "set_starting_thresholds"), cache.extra = file.info("resampling_functions.R")$mtime}
source("resampling_functions.R")
original_segcopy_data <- round(old_signal_matrix)
events_table <- as_tibble(rbind(2, original_segcopy_data)) %>%
  mutate_all(function(x) {ifelse(x == 2, 0, 1)})
unique_clones <- events_table %>%
  group_by_all() %>% summarise(num.cells = n())
print(paste(0, NA, nrow(unique_clones), NA, sum(events_table)))
naive_tree <- get_tree_from_data_matrix(original_segcopy_data, plot = T)
apply(old_signal_matrix, 2, range)
```

## Resampling tree

```{r resampling_tree, dependson=c("set_seed", "define_cnvs", "get_signal_matrix", "set_starting_thresholds"), cache.extra = file.info("resampling_functions.R")$mtime}
best_sample_data <- resample_matrix(
  old_signal_matrix, starting_thresholds = starting_thresholds,
  cns, num_loops = 10, num_samples = 1000, num_best = 100,
  plot_intermediate_trees = F, use_num_events = F)
best_sample_thresholds <- attr(best_sample_data, "thresholds")
resampled_segcopy_data <- round(apply(best_sample_data, 1:2, mean))
event_sizes <- cnv_table$end - cnv_table$start + 1
tree.arcs <- get_tree_from_data_matrix(
  resampled_segcopy_data, plot = T, event_sizes = event_sizes)

apply(best_sample_data, 1:2, mean) %>% as_tibble() %>% gather() %>%
  mutate(key = sub("V","", key)) %>%
  ggplot +
  # geom_violin(aes(x = key, y = value), bw = 0.1, col = "grey") +
  geom_dotplot(aes(x = key, y = value), binaxis = "y",
               stackdir = "center", binwidth = 0.01) +
  scale_x_discrete(limits = 1:ncol(resampled_segcopy_data))

apply(best_sample_data, 1:2, mean) %>% as_tibble() %>% gather() %>%
  mutate(key = sub("V","", key)) %>%
  ggplot +
  geom_violin(aes(x = key, y = value), bw = 0.1, col = "grey") +
  geom_dotplot(aes(x = key, y = value), binaxis = "y",
               stackdir = "center", binwidth = 0.01) +
  scale_x_discrete(limits = 1:ncol(resampled_segcopy_data))

colSums(!apply(best_sample_data, 1:2, function(x){all(x == x[1])}))
```

```{r, cache = F, eval = F}
# saveRDS(best_sample_data, "best_sample_data.rds")
# knitr::knit_exit()
best_sample_data <- readRDS("best_sample_data.rds")
resampled_segcopy_data <- round(apply(best_sample_data, 1:2, mean))
tree.arcs <- get_tree_from_data_matrix(resampled_segcopy_data, plot = T)
```

<!--
# QC

```{r no-compare_original_segcopy_data_to_final_data_matrix, results = "asis", cache.extra = file.info("resampling_chunks.R")$mtime}
```
-->

# CN events per clone {.tabset}

```{r genome_plots_with_clones, results = "asis", cache.extra = file.info("resampling_chunks.R")$mtime}
```

# Density Plots {.tabset}

```{r density_plots_post_tree, results = "asis", cache.extra = file.info("resampling_chunks.R")$mtime}
```

# Density Plot per clone {.tabset}

```{r density_plots_per_clone, results = "asis", cache.extra = file.info("resampling_chunks.R")$mtime}
```

# Sort overlapping events

```{r sort_overlapping_events}
# Merge 5 and 6, keep CN or 5 and end of 6
cnv_table[5, "end"] <- cnv_table[6, "end"]
cnv_table <- cnv_table[-6, ]
resampled_segcopy_data <- resampled_segcopy_data[, -6]
# Merge 7 and 9 (now 6 and 8)
cnv_table[6, "end"] <- cnv_table[8, "end"]
cnv_table <- cnv_table[-8, ]
resampled_segcopy_data <- resampled_segcopy_data[, -8]

event_sizes <- cnv_table$end - cnv_table$start + 1
tree.arcs <- get_tree_from_data_matrix(
  resampled_segcopy_data, plot = T, event_sizes = event_sizes)
```

# Table with all events

```{r}
cnv_table %>%
  mutate(length = end - start + 1) %>% 
  select(chr, start, end, length, n, num.cells) %>%
  DT::datatable(
      extensions = 'Buttons', options = list(
      dom = 'Bfrtip',
      buttons = list(
        'copy',
        list(extend = 'csv', filename = "Pt2_dia_CNAs"),
        list(extend = 'excel', filename = "Pt2_dia_CNAs"))
    )
  )
```

# COSMIC annotations {.tabset}

```{r no-cosmic_annotations, results = "asis", cache = F}
```

```{r echo = F, cache = F}
knitr::read_chunk("resampling_chunks.cosmic_annotations.R")
```

```{r cosmic_annotations_v90, results = "asis", cache = F}
```

# Annotated tree

```{r no-annotated_tree, fig.height = 8, fig.width = 8}
```

```{r echo = F, cache = F}
knitr::read_chunk("resampling_chunks.plot_final_trees.R")
tree_title = "Pt2 diagnosis"
```

```{r plot_final_trees, fig.height = 8, fig.width = 8, cache = FALSE}
```

## Tree with event matrix

```{r no-plot_final_trees_with_shared_events, fig.height = 11, fig.width = 11, cache = FALSE}
source("plot_cna_tree.R")

tree <- tree.arcs %>%
  select(parent, child, label) %>%
  rbind(tibble(parent = as.numeric(NA), child = 1, label = "")) %>%
  transmute(node_id = child, parent_id = parent, events = label) %>%
  arrange(node_id) %>%
  mutate(size = attr(tree.arcs, "counts")[, 1, drop = T])

datatable(cosmic_annotations, rownames = F)

event_labels <- cosmic_annotations %>%
  mutate(tree_label = case_when(
    str_length(genes) > 1 & str_length(genes) < 40 ~ genes,
    str_length(leukaemia_genes) > 1 & str_length(leukaemia_genes) < 40 ~
      paste0(leukaemia_genes, "/..."),
    str_length(leukaemia_genes) > 1 ~
      paste0(sub("/[^/]*$", "", substr(leukaemia_genes, 1, 41)), "/..."),
    str_length(genes) > 1 ~
      paste0(sub("/[^/]*$", "", substr(genes, 1, 41)), "/..."),
    TRUE ~ lesion
  )) %>% pull(tree_label)
event_labels <- paste(event_labels, ifelse(cnv_table$n < 2, "(loss)", "(gain)"), sep = " ")

plot_cna_tree(tree, event_labels = event_labels,
              x_left = 0, x_right = 41, y_top = 19, y_bottom = -27, title = tree_title)
```

## Small tree version

```{r no-plot_final_trees_small, fig.height = 6, fig.width = 6, cache = FALSE}
save.image("image.resampling_Pt2_dia_100k.RData")
load("image.resampling_Pt2_dia_100k.RData")
plot_cna_tree(tree, event_labels = event_labels, y_top = 19, title = tree_title, show_cna_matrix = F)
```


# Comparison to FISH data

```{r eval = F}
fish_clones_file <- "../data/fish/pt2.csv"

fish_clones <- read.csv(fish_clones_file, header = T)

kable(fish_clones)

fish_genes <- data.frame(gene = c("P16", "Pax5", "TEL", "AML"),
                         chr = c("chr9", "chr9", "chr12", "chr21"),
                         start = c(21967751, 36833272, 11802788, 36160098),
                         end = c(21995300, 37034103, 12048336, 37376965))
kable(fish_genes)

# New data frame to store the CN of the fish genes for each cell
cells <- colnames(segrenorm_data)[-c(1:3)]
cell_clones <- data.frame(matrix(NA, nrow = length(cells), ncol = nrow(fish_genes) + 1))
colnames(cell_clones) <- c("cell", as.character(fish_genes$gene))
rownames(cell_clones) <- cells
cell_clones$cell <- cells

for (cell in cells) {
  for (gene in 1:nrow(fish_genes)) {
    cn <- m.fit.cell.segcopy_data[
      as.character(m.fit.cell.segcopy_data$CHR) == as.character(fish_genes[gene, "chr"]) &
        m.fit.cell.segcopy_data$START <= fish_genes[gene, "end"] &
        m.fit.cell.segcopy_data$END >= fish_genes[gene, "start"], cell]
    if (length(unique(cn)) > 1) {
      if (all(cn < 2)) {
        cn <- min(cn)
      } else {
        cn <- max(cn)
      }
    }
    cell_clones[cell, as.character(fish_genes[gene, "gene"])] <- cn[1]
  }
}

```
