---
title: "Pt6 resampling analysis"
author: "Javier Herrero"
date: "`r date()`"
output:
  html_document:
    code_folding: hide
    toc: TRUE
    toc_float: TRUE
---

```{r}
knitr::read_chunk("resampling_chunks.R")
```

```{r setup, include=FALSE}
library(tidyverse)
library(cowplot)
library(ggExtra)
library(knitr)
library(optrees)
library(mclust)
library(DT)

knitr::opts_chunk$set(echo = TRUE)
opts_chunk$set(cache=TRUE)
opts_chunk$set(cache.lazy=FALSE)
opts_chunk$set(autodep = TRUE)

source("plots.R")
source("cnv_analysis.lib.R")
source("resampling_functions.R")
```

```{r set_seed}
set.seed(123456)
```

# Input data {.tabset}

```{r read_input, results = "asis"}
single_seg_renorm <- readRDS("cnv_copynumber_Pt6_dia_100k.single_renorm_diagnosticonly.RDS")
multi_seg_renorm <- readRDS("cnv_copynumber_Pt6_dia_100k.multi_seg_renorm_diagnosticonly.RDS")
library(xlsx)
meta_data <- read.xlsx(paste0("../data/Samples.xlsx"), sheetIndex = 1)
cells <- meta_data %>% filter(Day == "Diagnosis" & Patient == "Pt6") %>% pull(Cell) %>% as.character()
cells <- intersect(colnames(multi_seg_renorm[["05"]][, -c(1:3)]), cells)
multi_seg_renorm[["05"]] <- multi_seg_renorm[["05"]][, c("CHR", "START", "END", cells)]
single_seg_renorm[["10"]] <- single_seg_renorm[["10"]][, c("CHR", "START", "END", cells)]
```

There are `r length(cells)` cells in this dataset.

```{r define_cnvs, results = "asis"}
plotGenome(multi_seg_renorm[['05']])
plotGenome(single_seg_renorm[['10']])
plotFrequency(multi_seg_renorm[['05']])
plotFrequency(single_seg_renorm[['10']])
for (this_chr in paste0("chr", c(1:22, "X"))) {
  if (!is.null(opts_knit$get("output.dir"))) {
    cat(knit_child(text = paste0("\n## ", this_chr, "\n\n"), quiet = T))
  }
  plotChr(multi_seg_renorm[['05']], this_chr)
  plotChr(single_seg_renorm[['10']], this_chr)
}

get_cnv_table <- function(data) {
  cnv_table <- tibble()
  all_cells <- colnames(data)[-c(1:3)]
  chr_length_rle <- rle(data[, "CHR"])
  chr_starts <- cumsum(chr_length_rle$lengths) - chr_length_rle$lengths + 1
  names(chr_starts) <- chr_length_rle$values
  for (this_cell in all_cells) {
    for (this_chr in unique(data[, "CHR"])) {
      this_data <- data[data$CHR == this_chr, this_cell]
      rle <- rle(this_data)
      this_cnv_table <- tibble(
        from = cumsum(rle$lengths) - rle$lengths + chr_starts[this_chr],
        to = cumsum(rle$lengths) + chr_starts[this_chr] - 1,
        n = rle$values,
        num.cells = 1,
        cells = this_cell,
        chr = data[cumsum(rle$lengths) - rle$lengths + chr_starts[this_chr], "CHR"],
        start = data[cumsum(rle$lengths) - rle$lengths + chr_starts[this_chr], "START"],
        end = data[cumsum(rle$lengths) +  chr_starts[this_chr] - 1, "END"],
        n.bins = rle$lengths
        )
      cnv_table <- rbind(cnv_table, this_cnv_table %>% filter(n != 2))
    }
  }
  cnv_table <- cnv_table %>%
    group_by(from, to, n, chr, start, end, n.bins) %>%
    summarise(num.cells = n(), cells = paste(cells, collapse = ","))
  return(cnv_table)
}

curated_cnv_table <- read_csv("curated_cnv_table.Pt6_dia.csv")

cnv_table <- as.tibble(curated_cnv_table) %>% filter(keep == TRUE)
```

```{r get_signal_matrix}
data_env <- new.env()
load("data_Pt6_dia_100k_daignosticonly.rda", envir = data_env)
segrenorm_data <- data_env$segrenorm_data
rm(data_env)
segrenorm_data <- segrenorm_data[, c("CHR", "START", "END", cells)]

# ## Adjust for ploidy:
# # 1. read ploidy data (from Ginkgo)
# ploidy <- read.table("../data/100k/ploidy.txt", header = T, row.names = 1)
# # 2. Multiply signal (ca.1 centered) by ploidy
# segrenorm_data[, -c(1:3)] <- sweep(segrenorm_data[, -c(1:3)], 2,
#                                    ploidy[colnames(segrenorm_data)[-c(1:3)], ], "*")
# # 2. Divide signal by 2 (to make it 1 centered again, after ploidy adjustment)
# segrenorm_data[, -c(1:3)] <- sweep(segrenorm_data[, -c(1:3)], 2, 2, "/")

old_signal_matrix <- matrix(0, nrow = ncol(segrenorm_data) - 3, ncol = nrow(cnv_table))
colnames(old_signal_matrix) <- paste0("V", 1:ncol(old_signal_matrix))
for (i in 1:nrow(cnv_table)) {
  this_cnv <- cnv_table[i, , drop = T]

  old_signal_matrix[, i] <- segrenorm_data %>%
    filter(CHR == this_cnv$chr & START >= this_cnv$start & END <= this_cnv$end) %>%
    select(-c(1:3)) %>%
    colMeans() * 2
}
cns <- cnv_table$n
```


# Curated CN events before resampling {.tabset}

```{r set_starting_thresholds, dependson = c("get_signal_matrix")}
starting_thresholds <- list(
  mean = cnv_table$threshold,
  sd = cnv_table$sd)
```

```{r, cache = F}
DT::datatable(cnv_table)
```

```{r no-initial_density_plots, results = "asis", cache.extra = file.info("resampling_chunks.R")$mtime, dependson=c("define_cnvs", "get_signal_matrix", "set_starting_thresholds")}
```

```{r cache=F}
# knitr::knit_exit()
```


<!--
# Scatter plot

```{r no-pairwise_scatterplots, results = "asis", cache.extra = file.info("resampling_chunks.R")$mtime, dependson=c("define_cnvs", "get_signal_matrix", "set_starting_thresholds")}
```
-->

```{r cache=F}
# knitr::knit_exit()
```


# Reconstruct the tree

## Naive tree

```{r naive_tree, dependson=c("set_seed", "define_cnvs", "get_signal_matrix", "set_starting_thresholds"), cache.extra = file.info("resampling_functions.R")$mtime}
source("resampling_functions.R")
original_segcopy_data <- round(old_signal_matrix)
events_table <- as_tibble(rbind(2, original_segcopy_data)) %>%
  mutate_all(function(x) {ifelse(x == 2, 0, 1)})
unique_clones <- events_table %>%
  group_by_all() %>% summarise(num.cells = n())
print(paste(0, NA, nrow(unique_clones), NA, sum(events_table)))
naive_tree <- get_tree_from_data_matrix(original_segcopy_data, plot = T)
apply(old_signal_matrix, 2, range)
```

## Resampling tree

```{r resampling_tree, dependson=c("set_seed", "define_cnvs", "get_signal_matrix", "set_starting_thresholds"), cache.extra = file.info("resampling_functions.R")$mtime}
best_sample_data <- resample_matrix(
  old_signal_matrix, starting_thresholds = starting_thresholds,
  cns, num_loops = 10, num_samples = 1000, num_best = 100,
  plot_intermediate_trees = F, use_num_events = F)
best_sample_thresholds <- attr(best_sample_data, "thresholds")
resampled_segcopy_data <- round(apply(best_sample_data, 1:2, mean))
event_sizes <- cnv_table$end - cnv_table$start + 1
tree.arcs <- get_tree_from_data_matrix(
  resampled_segcopy_data, plot = T, event_sizes = event_sizes)

apply(best_sample_data, 1:2, mean) %>% as_tibble() %>% gather() %>%
  mutate(key = sub("V","", key)) %>%
  ggplot +
  # geom_violin(aes(x = key, y = value), bw = 0.1, col = "grey") +
  geom_dotplot(aes(x = key, y = value), binaxis = "y",
               stackdir = "center", binwidth = 0.01) +
  scale_x_discrete(limits = 1:ncol(resampled_segcopy_data))

apply(best_sample_data, 1:2, mean) %>% as_tibble() %>% gather() %>%
  mutate(key = sub("V","", key)) %>%
  ggplot +
  geom_violin(aes(x = key, y = value), bw = 0.1, col = "grey") +
  geom_dotplot(aes(x = key, y = value), binaxis = "y",
               stackdir = "center", binwidth = 0.01) +
  scale_x_discrete(limits = 1:ncol(resampled_segcopy_data))

colSums(!apply(best_sample_data, 1:2, function(x){all(x == x[1])}))
```

```{r, cache = F, eval = F}
# saveRDS(best_sample_data, "best_sample_data.rds")
# knitr::knit_exit()
best_sample_data <- readRDS("best_sample_data.rds")
resampled_segcopy_data <- round(apply(best_sample_data, 1:2, mean))
tree.arcs <- get_tree_from_data_matrix(resampled_segcopy_data, plot = T)
```

<!--
# QC

```{r no-compare_original_segcopy_data_to_final_data_matrix, results = "asis", cache.extra = file.info("resampling_chunks.R")$mtime}
```
-->

# CN events per clone {.tabset}

```{r genome_plots_with_clones, results = "asis", cache.extra = file.info("resampling_chunks.R")$mtime}
```

# Density Plots {.tabset}

```{r density_plots_post_tree, results = "asis", cache.extra = file.info("resampling_chunks.R")$mtime}
```

# Density Plot per clone {.tabset}

```{r density_plots_per_clone, results = "asis", cache.extra = file.info("resampling_chunks.R")$mtime}
```

# Fix annotations

Set chrX coordinates to 0-0 to mark this affects the whole chromosome

```{r fix_annotations}
# All cells with chr3 amplification happen to be hyperdiploid
cnv_table[7, c("start", "end")] <- 0
```

# COSMIC annotations {.tabset}

```{r no-cosmic_annotations, results = "asis", cache = F}
```

```{r echo = F, cache = F}
knitr::read_chunk("resampling_chunks.cosmic_annotations.R")
```

```{r cosmic_annotations_v90, results = "asis", cache = F}
```

# Annotated tree

```{r no-annotated_tree, fig.height = 8, fig.width = 8}
```

```{r echo = F, cache = F}
knitr::read_chunk("resampling_chunks.plot_final_trees.R")
tree_title = "Pt6 diagnosis"
```

```{r plot_final_trees, fig.height = 8, fig.width = 8, cache = FALSE}
```

## Tree with event matrix

```{r no-plot_final_trees_with_shared_events, fig.height = 11, fig.width = 11, cache = FALSE}
source("plot_cna_tree.R")

tree <- tree.arcs %>%
  select(parent, child, label) %>%
  rbind(tibble(parent = as.numeric(NA), child = 1, label = "")) %>%
  transmute(node_id = child, parent_id = parent, events = label) %>%
  arrange(node_id) %>%
  mutate(size = attr(tree.arcs, "counts")[, 1, drop = T])

datatable(cosmic_annotations, rownames = F)

event_labels <- cosmic_annotations %>%
  mutate(tree_label = case_when(
    str_length(genes) > 1 & str_length(genes) < 40 ~ genes,
    str_length(leukaemia_genes) > 1 & str_length(leukaemia_genes) < 40 ~
      paste0(leukaemia_genes, "/..."),
    str_length(leukaemia_genes) > 1 ~
      paste0(sub("/[^/]*$", "", substr(leukaemia_genes, 1, 41)), "/..."),
    str_length(genes) > 1 ~
      paste0(sub("/[^/]*$", "", substr(genes, 1, 41)), "/..."),
    TRUE ~ lesion
  )) %>% pull(tree_label)
event_labels <- paste(event_labels, ifelse(cnv_table$n < 2, "(loss)", "(gain)"), sep = " ")

plot_cna_tree(tree, event_labels = event_labels,
              x_left = 0, x_right = 41, y_top = 17, y_bottom = -29, title = tree_title)
```

## Small tree version

```{r no-plot_final_trees_small, fig.height = 6, fig.width = 6, cache = FALSE}
save.image("image.resampling_Pt6_dia_100k.RData")
load("image.resampling_Pt6_dia_100k.RData")
plot_cna_tree(tree, event_labels = event_labels, y_top = 19, title = tree_title, show_cna_matrix = F)
```
